Storm surge modelling in the Strait of Georgia

### 2014-11-26 RoughWork/manuscript.tex
Small edits

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). It is on average 220 km in length and 28 km wide and its maximum depth is 420 m \citep{thomson1981oceanography}. At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing. Many rivers feed fresh water into the Salish Sea system, the most significant source being from the Fraser River. Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal. This system is home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

%my attempt at physical oceanography in SoG- Susan feel free to edit.
Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months \citep{thomson1981oceanography}. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation \citep{masson2004modelling}. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model. 
%were Crean/Stronach modelled phases/amps too high or low?

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. Another group of modellers studying a larger domain noted high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. A finite volume model on an unstrcutured grid by \citet{yang2010multi} used a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semidiurnal amplitudes were generally too low. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling of this region.

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damaged caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  Additionally, there is a small contribution to increased sea level due to the thermal expansion of water during warmer years set by the El Nino Southern Oscillation (ENSO) \citep{abeys2011extreme}. %discuss this as a Kelvin wave

@@

%outline
This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep.  This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, the sea surface height anomaly from Neah Bay will be used in future real time water level predictions during the storm surge season. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.

>>>>

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). It is on average 220 km in length and 28 km wide and its maximum depth is 420 m \citep{thomson1981oceanography}. At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing. Many rivers feed fresh water into the Salish Sea system, the most significant source being from the Fraser River. Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

%my attempt at physical oceanography in SoG- Susan feel free to edit.
Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source of fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months \citep{thomson1981oceanography}. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation \citep{masson2004modelling}. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, set a fairly restrictive time step and impose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction. thei M$_2$ amplitudes were too high in the Strait of Juan de Fuca and too low in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. A finite volume model on an unstrcutured grid by \citet{yang2010multi} used a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semidiurnal amplitudes were generally too low. An updated version of this model produced high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered, notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling of this region.

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damaged caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There can also be a small contribution to increased sea level due to the thermal expansion of water during warmer years set by the El Nino Southern Oscillation (ENSO) \citep{abeys2011extreme}. %discuss this as a Kelvin wave

@@

%outline
This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain including a description of the ocean model we have employed is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, the sea surface height anomaly from Neah Bay will be used in future real-time water level predictions during the storm surge season. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.
